Explicit Jacobian matrix formulas for entropy stable summation-by-parts schemes

### 2020-03-02 docs/ESDG_jacobian/main.tex
updating jacobian paper

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{myproof}[1][\proofname]{%
  \proof Proof of #1%
}{\endproof}

@@

%% d in integrand
\newcommand*\diff[1]{\mathop{}\!{\mathrm{d}#1}}
\newcommand{\diag}[1]{{\rm diag}\LRp{#1}}
\newcommand{\td}[2]{\frac{{\rm d}#1}{{\rm d}{\rm #2}}}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\nor}[1]{\left\| #1 \right\|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRb}[1]{\left| #1 \right|}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\LRceil}[1]{\left\lceil #1 \right\rceil}
\newcommand{\LRl}[1]{\left. #1 \right|}
\newcommand{\pdn}[3]{\frac{\partial^{#3}#1}{\partial#2^{#3}}}
\newcommand{\Grad} {\ensuremath{\nabla}}
\newcommand{\jump}[1] {\ensuremath{\llbracket#1\rrbracket}}
\newcommand{\avg}[1] {\ensuremath{\LRc{\!\LRc{#1}\!}}}

@@

\subsection{Systems of conservation laws: vector-valued fluxes}
\label{sec:systems}

It is straightforward to extend the scheme from scalar nonlinear conservation laws to an $n\times n$ system of conservation laws.  Let $\bm{f}_{S}(\bm{u}_L,\bm{u}_R): \mathbb{R}^{n}\times \mathbb{R}^n \rightarrow \mathbb{R}^n$ denote an entropy conservative flux function for a 1D system of conservation laws.  We can modify the arrays and matrices in (\ref{eq:sbpmat}) to account for this.

Let $\bm{u}_{\Omega}$ now denote a vector of vectors 
\[
\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
\bm{u}_2\\
\vdots\\
\bm{u}_K}, \qquad \bm{u}_k = \bmat{
\bm{u}_{k,1}\\
\bm{u}_{k,2}\\
\vdots\\
\bm{u}_{k,N_p}
}, \qquad
\bm{u}_{k,j} = \bmat{
\bm{u}_{k,j,1}\\
\bm{u}_{k,j,2}\\
\vdots\\
\bm{u}_{k,j,n}
}
\]
for $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{k,j,i}$ denotes the $j$th degree of freedom for the $i$th component of the solution on the $k$th element.   Then, we can define global mass and differentiation matrices $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ as in (\ref{eq:sbpmat}), with the local blocks $\bm{M}, \bm{Q},\bm{B}_L, \bm{B}_R$ replaced by 
\eq{
\bm{M} &\longrightarrow \bm{M} \otimes \bm{I}_n, \qquad 
\bm{Q} \longrightarrow \bm{Q} \otimes \bm{I}_n,\\
\bm{B}_L &\longrightarrow \bm{B}_L \otimes \bm{I}_n, \qquad 
\bm{B}_R \longrightarrow \bm{B}_R \otimes \bm{I}_n,
}
where $\bm{I}_n$ is the $n\times n$ identity matrix.  Note that $\bm{Q}_{\Omega}$ as defined in (\ref{eq:sbpmat}) remains skew-symmetric under this new definition.  We also redefine the flux matrix to account for the vector-valued nature of the flux function $\bm{f}_S$ 
\eqlab{
\bm{F} = \bmat{
\bm{F}_{11} & \ldots & \bm{F}_{1K}\\
\vdots & \ddots & \vdots\\
\bm{F}_{K1} & \ldots & \bm{F}_{KK}
}, \qquad \LRp{\bm{F}_{k_1,k_2}}_{ij} = \diag{\bm{f}_S(\bm{u}_{k_1,i}, \bm{u}_{k_2,j})}.\label{eq:fluxsystem}
}
%\note{Redefine $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ using Kronecker product blocks. }
Then, an entropy conservative scheme is given by
\[
\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.
\]
We note that when computing Jacobians for systems of nonlinear conservation laws $\pd{\bm{f}_S}{\bm{y}}$, the derivative of the flux function is computed with respect to each component of the solution.

@@

We now assume that the global solution $\bm{u}(\bm{x},t) \in \mathbb{R}^n$ is vector-valued, and redefine local matrices so that they act on the global vector of vector-valued coefficients
\eq{
\bm{V} &\longrightarrow \bm{I}_K \otimes \LRp{\bm{V} \otimes \bm{I}_n}, \qquad \bm{P} \longrightarrow \bm{I}_K \otimes \LRp{\bm{P}  \otimes \bm{I}_n}\\ 
\bm{E} &\longrightarrow \bm{I}_K \otimes \LRp{\bm{E}  \otimes \bm{I}_n}, \qquad \bm{V}_h \longrightarrow \bm{I}_K \otimes \LRp{\bm{V}_h  \otimes \bm{I}_n}\\
\bm{Q} &\longrightarrow \bm{I}_K \otimes \LRp{\bm{Q}  \otimes \bm{I}_n}, \qquad \bm{B} \longrightarrow \bm{I}_K \otimes \LRp{\bm{B}  \otimes \bm{I}_n}\\
\bm{B}_L &\longrightarrow \bm{I}_K \otimes \LRp{\bm{B}_L  \otimes \bm{I}_n}, \qquad \bm{B}_R \longrightarrow \bm{I}_K \otimes \LRp{\bm{B}_R  \otimes \bm{I}_n}.
}
We can also use $\bm{Q}_h$ to construct a globally skew-symmetric differentiation matrix (see also \cite{chan2018efficient}).  Define the matrix $\bm{S} = \bm{Q}-\bm{Q}^T$, and define $\bm{Q}_{\Omega}$ as the global block matrix
\[
\bm{Q}_{\Omega} = \frac{1}{2}\LRs{\begin{array}{cc|cc|cc|cc}
\bm{S} &  \bm{E}^T\bm{B} & &&& && \\
 - \bm{B}\bm{E} &  & & \bm{B}_R && && -\bm{B}_L\\ [.2em]
 \hline\\[-1.0em]
&& \bm{S} &  \bm{E}^T\bm{B}&&&& \\
& -\bm{B}_L &  - \bm{B}\bm{E} &&& \bm{B}_R&&\\
 \hline\\[-1.0em]
&&&& \ddots & \ddots&  \\
&&&   -\bm{B}_L & \ddots&\ddots&   \ddots&\\
 \hline\\[-1.0em]
&&&&&\ddots& \bm{S} &  \bm{E}^T\bm{B} \\
\bm{B}_R &&& &&-\bm{B}_L &  - \bm{B}\bm{E} &\\
\end{array}},
\]

It was shown in \cite{parsani2016entropy, chan2017discretely} that, when either the mass matrix is non-diagonal or the nodal set does not contain appropriate boundary points, it is necessary to perform an entropy projection (or extrapolation \cite{chenreview}) step to ensure discrete entropy stability.  Let $\bm{v}(\bm{u})$ denote the entropy variables as a function of the conservative variables, and let $\bm{u}(\bm{v})$ denote the inverse mapping.  We define the entropy projected variables $\tilde{\bm{u}}_{\Omega}$ as
\[
\tilde{\bm{u}}_{\Omega} = \bm{u}\LRp{\bm{V}_h\bm{P}\bm{v}\LRp{\bm{V}\hat{\bm{u}}_{\Omega}}}.
\]
We redefine the flux block $\bm{F}_{k_1,k_2}$ in (\ref{eq:fluxsystem}) using the entropy projected variables
\[
\LRp{\bm{F}_{k_1,k_2}}_{ij} = \diag{\bm{f}_S(\tilde{\bm{u}}_{k_1,i}, \tilde{\bm{u}}_{k_2,j})}
\]
Then, an entropy conservative method is given by 
\[
\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\bm{V}^T\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.  
\]

@@

Redefine the nonlinear term 
\[
\bm{f}(\hat{\bm{u}}) = 2\bm{V}^T\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1}.
\]
We introduce $\pd{\bm{u}}{\bm{v}}$ and $\pd{\bm{v}}{\bm{u}}$ as Jacobians of the conservative variables with respect to the entropy variables and vice versa.  These have been explicitly derived for several equations (for example, the Jacobians for the compressible Navier-Stokes equations are given in \cite{hughes1986new}).

We can compute the Jacobian of $\bm{f}(\hat{\bm{u}})$ via the chain rule, and illustrate this for an entry $i\neq j$ of the Jacobian
\[
\LRp{\pd{\bm{f}}{\hat{\bm{u}}_{\Omega}}}_{ij} =  2 \bm{V}_h^T \pd{}{\LRp{\hat{\bm{u}}_{\Omega}}_j} \LRp{\LRp{\bm{Q}_h\circ\bm{F}}\bm{1}}_i.
\]
We focus on the latter term $\pd{}{\hat{\bm{u}}_{\Omega}} \LRp{\bm{Q}_h\circ\bm{F}}\bm{1}$
\eq{
\LRp{\pd{}{\hat{\bm{u}}_{\Omega}} \LRp{\bm{Q}_h\circ\bm{F}}\bm{1}}_{ij} &= \pd{}{\hat{\bm{u}}_{\Omega,j}} \sum_{k} \LRp{\bm{Q}_h}_{ik} \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_k} \\
&=  2 \bm{V}_h^T \sum_{k} \LRp{\bm{Q}_h}_{ik} \LRl{\pd{\bm{f}_S}{y}}_{\tilde{\bm{u}}_i,\tilde{\bm{u}}_k} \pd{\tilde{\bm{u}}_i}{\hat{\bm{u}}_{\Omega,j}}
}
We define the matrix $\pd{\tilde{\bm{f}}}{\tilde{\bm{u}}}$ as
\[
\LRp{\pd{\tilde{\bm{f}}}{\tilde{\bm{u}}}}_{ij} = \LRp{\bm{Q}_h}_{ij} \LRl{\pd{\bm{f}_S}{y}}_{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j} = \LRp{\bm{Q}_h\circ \bm{F}_y}_{ij}.
\]
Let $\tilde{\bm{v}} = \bm{V}_h\bm{P}\bm{v}\LRp{\bm{V}\hat{\bm{u}}_{\Omega}}$ denote the projected entropy variables at evaluated at volume quadrature points.
The vector $\pd{\tilde{\bm{u}}}{\hat{\bm{u}}_{\Omega}}$  can be further expanded as 
\[
\pd{\tilde{\bm{u}}}{\hat{\bm{u}}_{\Omega}} = \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}} \bm{V}_h \bm{P} \LRl{\pd{\bm{v}}{\bm{u}}}_{\tilde{\bm{v}}} \bm{V}.
\]
where the Jacobian matrices for the maps between conservative and entropy variables are block diagonal matrices given by
\[
\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}} = \bmat{
\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_1} &&\\
& \ddots &\\
&& \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_K}
}, \qquad \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_k} = \bmat{
\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_{1,k}} &&\\
 &\ddots &\\
&& \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_{N_p,k}}
},
\]
and the local block $\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_{j,k}}$ is simply the Jacobian $\pd{\bm{u}}{\bm{v}}$ evaluated at the $j$th nodal value $\tilde{\bm{u}}_{j,k}$ on the $k$th element.

Let $N_p, N_q, N_f$ denote the number of total basis functions, quadrature points, and face quadrature points respectively, and define $N_{\rm total} = N_q + N_f$.  The structure and dimensions of matrices involved in constructing the Jacobian matrix are illustrated as follows:
\[
\pd{\bm{f}}{\hat{\bm{u}}_{\Omega}} = \renewcommand\matscale{.67}
\raiserows{2}{\matbox{3}{7}{N_p}{N_{\rm total}}{\bm{V}_h^T}}
\raiserows{0}{\matbox{7}{7}{N_{\rm total}}{N_{\rm total}}{\pd{\tilde{\bm{f}}}{\tilde{\bm{u}}}}}
\raiserows{0}{\matbox{7}{7}{N_{\rm total}}{N_{\rm total}}{\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}}}}
\raiserows{0}{\matbox{7}{5}{N_{\rm total}}{N_q}{ \bm{V}_h\bm{P} }}
\raiserows{1}{\matbox{5}{5}{N_q}{N_q}{ \LRl{\pd{\bm{v}}{\bm{u}}}_{\tilde{\bm{v}}} }}
\raiserows{1}{\matbox{5}{3}{N_q}{N_p}{\bm{V}}}
\]

>>>>

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{myproof}[1][\proofname]{%
  \proof Proof of #1%
}{\endproof}
\newtheorem*{remark}{Remark}

@@

%% d in integrand
\newcommand*\diff[1]{\mathop{}\!{\mathrm{d}#1}}
\newcommand{\diag}[1]{{\rm diag}\LRp{#1}}
\newcommand{\td}[2]{\frac{{\rm d}#1}{{\rm d}{ {#2}}}}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\nor}[1]{\left\| #1 \right\|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRb}[1]{\left| #1 \right|}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\LRceil}[1]{\left\lceil #1 \right\rceil}
\newcommand{\LRl}[1]{\left. #1 \right|}
\newcommand{\pdn}[3]{\frac{\partial^{#3}#1}{\partial#2^{#3}}}
\newcommand{\Grad} {\ensuremath{\nabla}}
\newcommand{\jump}[1] {\ensuremath{\llbracket#1\rrbracket}}
\newcommand{\avg}[1] {\ensuremath{\LRc{\!\LRc{#1}\!}}}

@@

%\subsection{Dissipative terms}
%
%Entropy stable schemes are constructed by adding entropy dissipation to entropy conservative formulations.  We characterize such

\section{Systems of conservation laws}
\label{sec:systems}

In this section, we extend the explicit Jacobian formulas from scalar nonlinear conservation laws to an $n\times n$ system of conservation laws.  Let $\bm{f}_{S}(\bm{u}_L,\bm{u}_R): \mathbb{R}^{n}\times \mathbb{R}^n \rightarrow \mathbb{R}^n$ denote an entropy conservative flux function for a 1D system of conservation laws.  We first formulate a system of ODEs by modifying the definition of arrays and matrices in (\ref{eq:sbpmat}).

Let $\bm{u}_{\Omega}$ denote a vector of vectors 
\[
\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
\bm{u}_2\\
\vdots\\
\bm{u}_n}, \qquad \bm{u}_i = \bmat{
\bm{u}_{i,1}\\
\bm{u}_{i,2}\\
\vdots\\
\bm{u}_{i,K}
}, \qquad
\bm{u}_{i,k} = \bmat{
\bm{u}_{i,k,1}\\
\bm{u}_{i,k,2}\\
\vdots\\
\bm{u}_{i,k,N_p}
}
\]
for $\ell = 1,\ldots, n$, $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{\ell,k,j}$ denotes the $j$th degree of freedom for the $\ell$th component of the solution on the $k$th element.  
%We redefine these matrices as the $n \times n$ block matrices 
%\eq{
%&\bm{M}_{\Omega} \longrightarrow \bm{I}_n\otimes \bm{M}_{\Omega} = \bmat{\bm{M}_{\Omega} &  & \\
%& \ddots & \\
% &  & \bm{M}_{\Omega}},
%\\
%&\bm{Q}_{\Omega} \longrightarrow \bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} = 
%\bmat{\bm{Q}_{\Omega} & \ldots & \bm{Q}_{\Omega}\\
%\vdots & \ddots & \vdots\\
%\bm{Q}_{\Omega} & \ldots & \bm{Q}_{\Omega}},
%}
%where $\bm{e}_n\in \mathbb{R}^n$ is the vector of all ones, and $\bm{e}_n\bm{e}_n^T$ is the $n\times n$ matrix of all ones.  
Let $(k_1, j_1)$ and $(k_2, j_2)$ be tuples which correspond to row and column indices of a matrix.  
We now define the flux matrix $\bm{F}$ as the $n\times n$ block matrix containing flux interactions between different nodes
\eqlab{
\bm{F} = \bmat{
\bm{F}_1 &&\\
& \ddots &\\
&& \bm{F}_n
}, \qquad \LRp{\bm{F}_\ell}_{(k_1, j_1),(k_2,j_2)} = \LRp{\bm{f}_{S}\LRp{\bm{u}_{:, k_1,j_1},\bm{u}_{:, k_2,j_2}}}_{\ell}.
\label{eq:fluxsys}
}
where $\bm{u}_{:,k,j}$ denotes the vector containing all solution components at the $k$th element and $j$th node, and each block $\bm{F}_\ell$ for $\ell = 1,\ldots, n$ corresponds to the $\ell$th component of the vector-valued flux evaluated at solution states $\bm{u}_{:,k_1,j_1},\bm{u}_{:,k_2,j_2}$.

Let $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ denote the global mass and differentiation matrices in (\ref{eq:sbpmat}).  Then, an entropy conservative scheme is given by
\[
\LRp{\bm{I}_n\otimes \bm{M}_{\Omega}}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} } \circ \bm{F}}\bm{1} = \bm{0}.
\]
where $\bm{e}_n\in \mathbb{R}^n$ is the vector of all ones, and $\bm{e}_n\bm{e}_n^T$ is the $n\times n$ matrix of all ones.

We now construct explicit Jacobian matrix formulas for systems of nonlinear conservation laws.  The right hand side function $\bm{f}(\bm{u})$ for systems can be rewritten as
\[
\bm{f}(\bm{u}) = 2\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} 
= 2\bmat{(\bm{Q}_{\Omega}\circ \bm{F}_1) \\ 
\vdots\\
(\bm{Q}_{\Omega}\circ \bm{F}_n)
}\bm{1}
\]
Then, the Jacobian matrix is 
\eqlab{
\pd{\bm{f}}{\bm{u}} &= 2\bmat{
\bm{Q}_{\Omega}\circ \bm{F}_{1,\bm{u}_1} & \ldots & \bm{Q}_{\Omega}\circ \bm{F}_{1,\bm{u}_n}\\ 
\vdots & \ddots & \vdots\\
\bm{Q}_{\Omega}\circ \bm{F}_{n,\bm{u}_1} & \ldots & \bm{Q}_{\Omega}\circ \bm{F}_{n,\bm{u}_n}
} \label{eq:fluxjac}\\
&= \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega}} \circ 
\bmat{
\bm{F}_{1,\bm{u}_1} & \ldots &\bm{F}_{1,\bm{u}_{n}}\\
\vdots & \ddots & \vdots \\
\bm{F}_{n,\bm{u}_1} & \ldots &\bm{F}_{n,\bm{u}_n}\nonumber
}.
}
Here, the flux matrices $\bm{F}_{i,\bm{u}_j}$ are evaluated via
\[
\LRp{\bm{F}_{i,\bm{u}_j}}_{j_1k_1,j_2k_2} = \LRl{\pd{\LRp{\bm{f}_S}_i}{\bm{u}_j}}_{\bm{u}_{:,k_1,j_1},\bm{u}_{:,k_2,j_2}},
\]
where $\pd{\LRp{\bm{f}_S}_i}{\bm{u}_{j}}$ denotes the derivative of the $i$th component of the flux $\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R}$ with respect to the $j$th solution component of $\bm{u}_R$.  Thus, each block $\bm{F}_{i,\bm{u}_j}$ corresponds to an entry of the Jacobian of $\bm{f}_S(\bm{u}_L,\bm{u}_R)$).

\begin{remark}
We note that an alternative ordering of degrees of freedom \cite{crean2018entropy} yields a Jacobian matrix with a more compact bandwidth.   A smaller matrix bandwidth can also be achieved by bandwidth-minimizing algorithms such as the reverse Cuthill-McKee algorithm \cite{cuthill1969reducing}.
%Let $\bm{u}_{\Omega}$ now denote a vector of vectors 
%\[
%\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
%\bm{u}_2\\
%\vdots\\
%\bm{u}_K}, \qquad \bm{u}_k = \bmat{
%\bm{u}_{k,1}\\
%\bm{u}_{k,2}\\
%\vdots\\
%\bm{u}_{k,N_p}
%}, \qquad
%\bm{u}_{k,j} = \bmat{
%\bm{u}_{k,j,1}\\
%\bm{u}_{k,j,2}\\
%\vdots\\
%\bm{u}_{k,j,n}
%}
%\]
%for $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{k,j,i}$ denotes the $j$th degree of freedom for the $i$th component of the solution on the $k$th element.   Define global mass and differentiation matrices $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ as in (\ref{eq:sbpmat}), with the local blocks $\bm{M}, \bm{Q},\bm{B}_L, \bm{B}_R$ replaced by 
%\eq{
%\bm{M} &\longrightarrow \bm{M} \otimes \bm{I}_n, \qquad 
%\bm{Q} \longrightarrow \bm{Q} \otimes \bm{I}_n,\\
%\bm{B}_L &\longrightarrow \bm{B}_L \otimes \bm{I}_n, \qquad 
%\bm{B}_R \longrightarrow \bm{B}_R \otimes \bm{I}_n,
%}
%where $\bm{I}_n$ is the $n\times n$ identity matrix.  Note that $\bm{Q}_{\Omega}$ as defined in (\ref{eq:sbpmat}) remains skew-symmetric under this new definition.  We also redefine the flux matrix to account for the vector-valued nature of the flux function $\bm{f}_S$ 
%\eqlab{
%\bm{F} = \bmat{
%\bm{F}_{11} & \ldots & \bm{F}_{1K}\\
%\vdots & \ddots & \vdots\\
%\bm{F}_{K1} & \ldots & \bm{F}_{KK}
%}, \qquad \LRp{\bm{F}_{k_1,k_2}}_{ij} = \diag{\bm{f}_S(\bm{u}_{k_1,i}, \bm{u}_{k_2,j})}.\label{eq:fluxsystem}
%}
%Then, an entropy conservative formulation is given by
%\[
%\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.
%\]
%Explicit Jacobians can also be constructed for this ordering.  We omit it for conciseness.
\end{remark}

%\subsection{Explicit Jacobian matrix formulas for systems of nonlinear conservation laws}

@@

We can also use $\bm{Q}_h$ to construct a globally skew-symmetric differentiation matrix (see also \cite{chan2018efficient}).  Define the matrix $\bm{S} = \bm{Q}-\bm{Q}^T$, and define $\bm{Q}_{\Omega}$ as the global block matrix
\[
\bm{Q}_{\Omega} = \frac{1}{2}\LRs{\begin{array}{cc|cc|cc|cc}
\bm{S} &  \bm{E}^T\bm{B} & &&& && \\
 - \bm{B}\bm{E} &  & & \bm{B}_R && && -\bm{B}_L\\ [.2em]
 \hline\\[-1.0em]
&& \bm{S} &  \bm{E}^T\bm{B}&&&& \\
& -\bm{B}_L &  - \bm{B}\bm{E} &&& \bm{B}_R&&\\
 \hline\\[-1.0em]
&&&& \ddots & \ddots&  \\
&&&   -\bm{B}_L & \ddots&\ddots&   \ddots&\\
 \hline\\[-1.0em]
&&&&&\ddots& \bm{S} &  \bm{E}^T\bm{B} \\
\bm{B}_R &&& &&-\bm{B}_L &  - \bm{B}\bm{E} &\\
\end{array}},
\]
We also construct global interpolation, projection, and extrapolation matrices
\eq{
\bm{V} &\longrightarrow \bm{I}_K \otimes \bm{V}, \qquad \bm{P} \longrightarrow \bm{I}_K \otimes \bm{P} \\ 
\bm{E} &\longrightarrow \bm{I}_K \otimes \bm{E}, \qquad \bm{V}_h \longrightarrow \bm{I}_K \otimes \bm{V}_h
}
Finally, we assume that the global solution $\bm{u}(\bm{x},t) \in \mathbb{R}^n$ is vector-valued, and order the solution coefficients as in Section~\ref{sec:systems}.

It was shown in \cite{parsani2016entropy, chan2017discretely} that, when either the mass matrix is non-diagonal or the nodal set does not contain appropriate boundary points, it is necessary to perform an entropy projection (or extrapolation \cite{chenreview}) step to ensure discrete entropy stability.  Let $\bm{v}(\bm{u})$ denote the entropy variables as a function of the conservative variables, and let $\bm{u}(\bm{v})$ denote the inverse mapping.  We define the entropy projected variables $\tilde{\bm{u}}_{\Omega}$ as
\[
\tilde{\bm{u}}_{\Omega} = \bm{u}\LRp{\bm{V}_h\bm{P}\bm{v}\LRp{\bm{V}\hat{\bm{u}}_{\Omega}}}.
\]
We evaluate each flux block $\bm{F}_{\ell}$ in (\ref{eq:fluxsys}) using the entropy projected variables
\[
\LRp{\bm{F}_\ell}_{(k_1, j_1),(k_2,j_2)} = \LRp{\bm{f}_{S}\LRp{\tilde{\bm{u}}_{:, k_1,j_1},\tilde{\bm{u}}_{:, k_2,j_2}}}_{\ell}.
%\LRp{\bm{F}_{i,\bm{u}_j}}_{j_1k_1,j_2k_2} = \LRl{\pd{\LRp{\bm{f}_S}_i}{\bm{u}_j}}_{\tilde{\bm{u}}_{:,k_1,j_1},\tilde{\bm{u}}_{:,k_2,j_2}},
\]
Then, an entropy conservative method is given by 
\[
\LRp{\bm{I}_n\otimes \bm{M}_{\Omega}}\td{\bm{u}_{\Omega}}{t} + 2\LRp{\bm{I}_n\otimes \bm{V}_h}^T\LRp{ \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} } \circ \bm{F}}\bm{1} = \bm{0}.
%\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\bm{V}^T\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.  
\]

@@

Redefine the nonlinear term 
\[
\bm{f}(\hat{\bm{u}}) = 2\LRp{\bm{I}_n\otimes \bm{V}_h}^T\LRp{ \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega}} \circ \bm{F}}\bm{1}.
\]
We introduce $\pd{\bm{u}}{\bm{v}}$ and $\pd{\bm{v}}{\bm{u}}$ as Jacobians of the conservative variables with respect to the entropy variables and vice versa.  These have been explicitly derived for several equations (for example, the Jacobians for the compressible Navier-Stokes equations are given in \cite{hughes1986new}).

We can compute the Jacobian of $\bm{f}(\hat{\bm{u}})$ via the chain rule.  We assume $n=1$ for simplicity, and illustrate this for an entry $i\neq j$ of the Jacobian
\[
\LRp{\pd{\bm{f}}{\hat{\bm{u}}_{\Omega}}}_{ij} =  2 \bm{V}_h^T \pd{}{\LRp{\hat{\bm{u}}_{\Omega}}_j} \LRp{\LRp{\bm{Q}_{\Omega}\circ\bm{F}}\bm{1}}_i.
\]
We focus on the latter term $\pd{}{\hat{\bm{u}}_{\Omega}} \LRp{\bm{Q}_{\Omega}\circ\bm{F}}\bm{1}$
\eq{
\LRp{\pd{}{\hat{\bm{u}}_{\Omega}} \LRp{\bm{Q}_{\Omega}\circ\bm{F}}\bm{1}}_{ij} &= \pd{}{\hat{\bm{u}}_{\Omega,j}} \sum_{k} \LRp{\bm{Q}_{\Omega}}_{ik} \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_k} \\
&=  2 \bm{V}_h^T \sum_{k} \LRp{\bm{Q}_{\Omega}}_{ik} \LRl{\pd{\bm{f}_S}{y}}_{\tilde{\bm{u}}_i,\tilde{\bm{u}}_k} \pd{\tilde{\bm{u}}_i}{\hat{\bm{u}}_{\Omega,j}}
}
We observe that the term $\pd{\tilde{\bm{u}}_i}{\hat{\bm{u}}_{\Omega,j}}$ does not disappear as it did in the proof of Lemma~\ref{lemma:explicitJ}.  We thus treat the Jacobian matrix in two parts.  First, we define the matrix $\pd{\tilde{\bm{f}}}{\tilde{\bm{u}}}$ as
\[
\LRp{\pd{\tilde{\bm{f}}}{\tilde{\bm{u}}}}_{ij} = \LRp{\bm{Q}_{\Omega}}_{ij} \LRl{\pd{\bm{f}_S}{y}}_{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j} = \LRp{\bm{Q}_{\Omega}\circ \bm{F}_y}_{ij}.
\]
Now, let $\tilde{\bm{v}}$ denote the projected entropy variables at evaluated at volume quadrature points
\[
\tilde{\bm{v}} = \bm{V}_h\bm{P}\bm{v}\LRp{\bm{V}\hat{\bm{u}}_{\Omega}}.
\]
The vector $\pd{\tilde{\bm{u}}}{\hat{\bm{u}}_{\Omega}}$  can be further expanded as 
\[
\pd{\tilde{\bm{u}}}{\hat{\bm{u}}_{\Omega}} = \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}} \bm{V}_h \bm{P} \LRl{\pd{\bm{v}}{\bm{u}}}_{\tilde{\bm{v}}} \bm{V}.
\]
where the Jacobian matrices for the maps between conservative and entropy variables are block diagonal matrices given by
\[
\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}} = \bmat{
\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_1} &&\\
& \ddots &\\
&& \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_K}
}, \qquad \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_k} = \bmat{
\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_{1,k}} &&\\
 &\ddots &\\
&& \LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_{N_p,k}}
},
\]
and the local block $\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}_{j,k}}$ is the Jacobian matrix $\pd{\bm{u}}{\bm{v}}$ evaluated at the $j$th nodal solution value $\tilde{\bm{u}}_{j,k}$ on the $k$th element.

Let $N_p, N_q, N_f$ denote the number of total basis functions, quadrature points, and face quadrature points respectively, and define $N_{\rm total} = N_q + N_f$.  The structure and dimensions of matrices involved in constructing the Jacobian matrix are illustrated as follows:
\[
\pd{\bm{f}}{\hat{\bm{u}}_{\Omega}} = \renewcommand\matscale{.67}
\raiserows{2}{\matbox{3}{7}{N_p}{N_{\rm total}}{\bm{V}_h^T}}
\raiserows{0}{\matbox{7}{7}{N_{\rm total}}{N_{\rm total}}{\bm{Q}_{\Omega}\circ\pd{\tilde{\bm{f}}}{\tilde{\bm{u}}}}}
\raiserows{0}{\matbox{7}{7}{N_{\rm total}}{N_{\rm total}}{\LRl{\pd{\bm{u}}{\bm{v}}}_{\tilde{\bm{u}}}}}
\raiserows{0}{\matbox{7}{5}{N_{\rm total}}{N_q}{ \bm{V}_h\bm{P} }}
\raiserows{1}{\matbox{5}{5}{N_q}{N_q}{ \LRl{\pd{\bm{v}}{\bm{u}}}_{\tilde{\bm{v}}} }}
\raiserows{1}{\matbox{5}{3}{N_q}{N_p}{\bm{V}}}
\]

We note that several matrices become simplified under common assumptions for entropy stable DG discretizations.  These typically fall into two categories: collocated volume nodes and collocated volume and surface nodes \cite{shadpey2019energy}.  For collocated volume nodes, the solution is represented using a nodal Lagrange basis constructed using $N_q = N_p$ volume quadrature nodes.  Both collocated volume and surface nodes refers to the case when both the volume nodes collocated and the volume quadrature nodes also include surface quadrature nodes as a subset \cite{gassner2013skew, chen2017entropy}.  For example, if only volume nodes are collocated \cite{chan2018efficient}, then $\bm{V} = \bm{I}$.  If both volume and surface nodes are collocated, the system reduces to the simplified system described in Section~\ref{sec:systems} using the fact that $\pd{\bm{u}}{\bm{v}} = \LRp{\pd{\bm{v}}{\bm{u}}}^{-1}$.
